---
title: "Multilevel Model Estimation Using the Lueven Clinical Data Set"
author:
  - "[Ginette Lafit](/presenters/ginette-lafit.md)"
date: today
toc: true
format:
    html: default
    pdf: default
execute:
  freeze: auto
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries

## Prelim - Installing libraries used in this script (whenever is necessary).

```{r chunk-install-package, echo=TRUE, warning=TRUE, results="hide", message=FALSE}
# This code chunk simply makes sure that all the
# libraries used here are installed.
# Check if R packages are installed
## Do not run because we do not want to install packages (this should be your decision)

#list.of.packages = c("tidyverse","data.table","psych","viridis","nlme")
#new.packages = list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
#if(length(new.packages)) install.packages(new.packages)

# Delete # if the package PowerLAPIM is not installed yet
#library(devtools)
#devtools::install_github("ginettelafit/PowerLAPIM", force = T)
```

### Upload libraries to be used in the analysis

```{r chunk-libraries, echo=TRUE, warning=TRUE, results="hide", message=FALSE}
library(tidyverse) # reshaped data, variable transformation, plots (using library ggplot2)
library(data.table) # to create lagged outcome
library(psych) # to compute descriptive statistics & some psychometrics
library(viridis) # color scales adapted for colorblindness
library(nlme) # to estimate mixed-effects models
set.seed(1234) # we set the random seed to reproduce the analysis
```

# The dataset

We use data from @heininga2019dynamical; this study applies the ESM methodology
to study emotion dynamics in people with Major Depressive Disorder. The study
consist of an ESM testing period of 7 days in which participants had to fill out
questions about mood and social context on their daily lives ten times a day
(i.e., 70 measurement occasions). The data set contains 38 participants
diagnosed with MDD and 40 control subjects. Participants filled out the ESM
questionnaires in a stratified random interval scheme between 9:30 AM and 9:30
PM.

The dataset contains the following variables: `PID` that denotes the individual
identification number, `day` is a variable that ranges from 1 to 7 and
identifies the day of ESM testing, `daybeep` is a variable that ranges from 1 to
10 and identifies the number of the prompt or beep within a day. `PA` is the
Positive Affect computed as the mean of items: 'How happy do you feel at the
moment?', 'How relaxed do you feel at the moment?' and 'How euphoric do you feel
at the moment?'. `NA.` is the Negative Affect computed as the mean of items:
'How depressed do you feel at the moment?', 'How stressed do you feel at the
moment?', 'How anxious do you feel at the moment?', 'How angry do you feel at
the moment?' and 'How restless do you feel at the moment?'. `anhedonia`
corresponds to the ESM item 'To what degree do you find it difficult to
experience pleasure in activities at the moment?'. `MDD` is a dummy variable
equal to one when the individual has been diagnosed with MDD and 0 otherwise,
finally `QIDS` denotes the sum of the items of the Quick Inventory of Depressive
Symptomatology (i.e. QIDS) [@rush200316]. QIDS was measured before the ESM
testing period.

First we are going to load the data set:

```{r chunk-load-data, echo = TRUE}
# Load data set
load(file="assets/data/clinical-dataset.RData")

# Find the dimensions
dim(data)

# Find the structure
str(data)

# See the first 6 rows
head(data)

# See the last 6 rows
tail(data)

# Find the column names
names(data)

# Summary of the data
summary(data)

# Number of participants
length(unique(data$PID))

# Add variable number of observation per person
data$obs = rep(0,nrow(data))
for (i in unique(data$PID)){
    data$obs[which(data$PID==i)] = 1:length(which(data$PID==i))
}

table(data$obs)
```

# Visualizations and descriptive statistics

We first compute descriptive statistics including number of participant, number
of observations per day, and compliance.

```{r, chunk-descriptive-statistics, echo=TRUE}
# Get number of participants
length(unique(data$PID))

# Obtain number of participants diagnosed with MDD
length(unique(data$PID[data$MDD==1]))

# Obtain number of participants in the control group
length(unique(data$PID[data$MDD==0]))

# Get the number of assessment per day
table(data$PID)
# Get the number of assessment per day for each participant
beeps.person = lapply(data$PID, function(i) table(data$day[which(data$PID==i)]))
# Show results for some of the participants
beeps.person[1:6]

# Compute a binary variable indicating if a participant answered a beep. We take the ESM item Happy as reference
data$Compliance = ifelse(is.na(data$PA)==FALSE, 1, 0)

# Mean, median of the compliance across all participants
describe(data$Compliance)

# Compliance per participant
data.compliance.person = aggregate(data$Compliance, by = list(data$PID), mean, na.rm = TRUE)

# See the first 6 rows
head(data.compliance.person)

# See the last 6 rows
tail(data.compliance.person)

# Obtain descriptive statistics of person's average compliance
describe(data.compliance.person$x)
```

Next, we obtain visualizations and descriptive statistics of the distribution of
the person-level or time-invariant variables.

```{r, chunk-descriptive-statistics-level2-variables, echo=TRUE}
# We create a variable including the variable diagnosis (1 = MDD, 0 = control), and depression (QIDS) for each participant
dt.person = aggregate(cbind(data$MDD,data$QIDS), by = list(data$PID), mean, na.rm = TRUE)
colnames(dt.person) = c("Group.1","MDD","QIDS")

# See the first 6 rows
head(dt.person)

# See the last 6 rows
tail(dt.person)

# Descriptive statistics for time-invariant variable QIDS
describe(dt.person$QIDS)

# Descriptive statistics for time-invariant variable QIDS for MDD=1
describe(dt.person$QIDS[dt.person$MDD==1])

# Descriptive statistics for time-invariant variable QIDS for MDD=0
describe(dt.person$QIDS[dt.person$MDD==0])
```

We now focus on time-varying variables, we obtain visualization and descriptive
statistics for the time-varying variable negative affect (NA).

```{r, chunk-descriptive-statistics-level1-variables, echo=TRUE}
#| fig-align: center

# Histogram for the time-varying variable negative affect (NA.)
ggplot(data, aes(NA.)) + geom_histogram(bins=30) +  scale_fill_viridis() + theme_bw()

# Histogram for the time-varying variable NA. by MDD
ggplot(data, aes(NA.)) + geom_histogram(bins=30) +
  facet_wrap(~MDD) + scale_fill_viridis() + theme_bw()

# Descriptive statistics for NA.
describe(data$NA.)

# Descriptive statistics for NA. for MDD=1
describe(data$NA.[data$MDD==1])

# Descriptive statistics for NA. for MDD=1
describe(data$NA.[data$MDD==0])

# Distribution of happy per participant
data.table.dt = setDT(na.omit(data))
data.table.dt[, as.list(summary(NA., na.omit = TRUE)), by = PID]

# We randomly select 10 participants for plotting the distribution of the time-varying variable NA.
n.ID.sample = sample(unique(data$PID),4)
data.person.sample = data[which(data$PID %in% n.ID.sample),]

# Histogram for the time-varying variable happy by person
ggplot(data.person.sample, aes(NA.)) + geom_histogram(color="black", fill="white",bins=30) +
  facet_wrap(~PID) +  scale_fill_viridis() + theme_bw()

# Plot the trajectories of the time-varying variable NA by person
data.person.sample %>%
  ggplot(aes(x = obs, y = NA.)) +
  geom_point() +
  geom_line() +  # add lines to connect the data for each person
  facet_wrap( ~ PID) +  scale_fill_viridis() + theme_bw()

# We create a variable including the MDD (1 = MDD, 0 = Controls), and person's means of the time-varying variable NA.
dt.person = aggregate(cbind(data$MDD,data$NA.), by = list(data$PID), mean, na.rm = TRUE)
colnames(dt.person) = c("Group.1","MDD","NA.")

# See the first 6 rows
head(dt.person)

# See the last 6 rows
tail(dt.person)

# Descriptive statistics for person's means of the time-varying variable NA.
describe(dt.person$NA.)

# Descriptive statistics for person's means of the time-varying variable NA. for MDD=1
describe(dt.person$NA.[dt.person$MDD==1])

# Descriptive statistics for person's means of the time-varying variable NA. for MDD=0
describe(dt.person$NA.[dt.person$MDD==0])

# We create a variable including the MDD (1 = MDD, 0 = Controls), and person's standard deviation of the time-varying variable NA.
dt.person.sd = aggregate(data$NA., by = list(data$PID,data$MDD), sd, na.rm = TRUE)
colnames(dt.person.sd) = c("Group.1","MDD","NA.")

# See the first 6 rows
head(dt.person.sd)

# See the last 6 rows
tail(dt.person.sd)

# Descriptive statistics for person's standard deviation of the time-varying variable NA.
describe(dt.person.sd$NA.)

# Descriptive statistics for person's standard deviation of the time-varying variable NA. for MDD=1
describe(dt.person.sd$NA.[dt.person.sd$MDD==1])

# Descriptive statistics for person's standard deviation of the time-varying variable NA. for MDD=0
describe(dt.person.sd$NA.[dt.person.sd$MDD==0])
```

# Example 1: estimate the effect of a continuous time-varying predictor

The first illustrative example shows how to estimate the effect of a
time-varying predictor on the outcome of interest. Considering the Leuven
clinical study, we are interested in studying the impact of Anhedonia on
Negative Affect in daily life on patients with major depressive disorder (MDD).

We use the data of 38 individuals diagnosed with MDD; we select the individuals
diagnosed with MDD.

```{r chunk-subset-data, echo = TRUE}
data.MDD = data[which(data$MDD==1),]
```

First, we are going to estimate the individual means, the mean across all participants, and the standard deviation of the variable `anhedonia`.

```{r chunk-centering-l1-predictor-1, echo = TRUE}
# Compute the group mean of anhedonia
groupmean_X = aggregate(data.MDD$anhedonia, list(data.MDD$PID), FUN = mean, data=data.MDD, na.rm=TRUE)
mean_X = mean(groupmean_X[,2])
mean_X

# Compute the standard deviation of anhedonia
sd_X = sd(data.MDD$anhedonia, na.rm=TRUE)
sd_X
```

Next, we are going to person mean-centered the variable `anhedonia`.

```{r chunk-centering-l1-predictor-2, echo = TRUE}
# Centered within individuals anhedonia.
N.i = unique(data.MDD$PID)
anhedonia.c = rep(0, nrow(data.MDD))

for(i in N.i){
    anhedonia.c[which(data.MDD$PID==i)] = data.MDD$anhedonia[which(data.MDD$PID==i)] -
    mean(data.MDD$anhedonia[which(data.MDD$PID==i)],na.rm=TRUE)
}

data.MDD = cbind(data.MDD,anhedonia.c)
```

We estimate the linear mixed-effects model assuming AR(1) errors:

```{r chunk-fit-model-1, echo = TRUE}
# fit a linear mixed-effects model to data
fit.Model.1 = lme(NA. ~ 1 + anhedonia.c, random = ~1 + anhedonia.c|PID,na.action=na.omit, data=data.MDD, correlation=corAR1(), method="REML")
```

The summary of the estimation results is given by:

```{r chunk-summary-model-1, echo = TRUE}
summary(fit.Model.1)
```

Obtain confidence intervals:

```{r chunk-summary-CI-model-1, echo = TRUE}
intervals(fit.Model.1, which = "fixed")
```

The estimated fixed intercept is given by:

```{r chunk-summary-model-1-FE-1, echo = TRUE}
# Extract fixed effect coefficients
# extract the value of fixed intercept
coef(summary(fit.Model.1))[1,1]
```

the effect of the level-2 continuous variable on the intercept is extracted as
follows:

```{r chunk-summary-model-1-FE-2, echo = TRUE}
# extract the value of the fixed slope
coef(summary(fit.Model.1))[2,1]
```

The standard deviation and autocorrelation of the level-1 residuals are
extracted as follows:

```{r chunk-summary-model-1-L1-Errors, echo = TRUE}
# Extract level-1 residuals standard deviation
as.numeric(VarCorr(fit.Model.1)[3,2])

# Extract level-1 residuals correlation between consecutive points
as.numeric(coef(fit.Model.1$modelStruct$corStruct,unconstrained=FALSE))
```

The standard deviation of the random intercept is given by:

```{r chunk-summary-model-1-RE-1, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random intercept
as.numeric(VarCorr(fit.Model.1)[1,2])
```

The standard deviation of the random slope is given by:

```{r chunk-summary-model-1-RE-2, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.1)[2,2])
```

The correlation between the random intercept and the random slope is given by:

```{r chunk-summary-model-1-RE-3, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.1)[2,3])
```

# Example 2: estimate cross-level interaction effect between a continuous time-varying predictor and a continuous time-invariant predictor

We now show how to estimate a cross-level interaction effect between a
continuous time-varying predictor and continuous time-invariant predictor. In
particular, we are interested in studying if depression moderated the impact of
anhedonia on negative affect in daily life for individuals diagnosed with MDD.

Before estimating the model, we are going to compute the mean and standard
deviation of the Level 2 variable `QIDS`.

```{r chunk-subset-data-stats-MDD, echo = TRUE}
# Compute the mean of W
groupmean_W = aggregate(data.MDD$QIDS, list(data.MDD$PID), FUN = mean, data=data.Controls, na.rm=TRUE, method = "REML")
mean_W = mean(groupmean_W[,2])
mean_W

# Compute the standard deviation of W

sd_W = sd(groupmean_W[,2])
sd_W
```

Next, we are going to mean centered the variable `QIDS` using the mean estimated
above:

```{r chunk-centering-l2-predictor-controls, echo = TRUE}
# Centered QIDS

N.i = unique(data.MDD$PID)
QIDS.c = rep(0,nrow(data.MDD))
for(i in N.i){
QIDS.c[which(data.MDD$PID==i)] = data.MDD$QIDS[which(data.MDD$PID==i)] - mean_W
}

data.MDD = cbind(data.MDD,QIDS.c)
```

Next, we estimate the linear mixed-effects model assuming AR(1) errors:

```{r chunk-fit-model-2, echo = TRUE}
# fit a linear mixed-effects model to data
fit.Model.2 = lme(NA. ~ 1 + anhedonia.c + anhedonia.c*QIDS.c, random = ~1 + anhedonia.c|PID,na.action=na.omit, data=data.MDD, correlation=corAR1(), method="REML")
```

The summary of the estimation results is given by:

```{r chunk-summary-model-2, echo = TRUE}
summary(fit.Model.2)
```

Obtain confidence intervals:

```{r chunk-summary-CI-model-2, echo = TRUE}
intervals(fit.Model.2, which = "fixed")
```

The estimated fixed intercept is given by:

```{r chunk-summary-model-2-FE, echo = TRUE}
# Extract fixed effect coefficients
# extract the value of fixed intercept
coef(summary(fit.Model.2))[1,1]
```

the effect of the level-2 continuous variable on the intercept is extracted as
follows:

```{r chunk-summary-model-2-FE-1, echo = TRUE}
# extract the value of the fixed slope
coef(summary(fit.Model.2))[2,1]
```

the effect of the level-2 continuous variable on the intercept is extracted as
follows:

```{r chunk-summary-model-2-FE-2, echo = TRUE}
# extract the value of the fixed slope
coef(summary(fit.Model.2))[3,1]
```

the effect of the level-2 continuous variable on the intercept is extracted as
follows:

```{r chunk-summary-model-2-FE-3, echo = TRUE}
# extract the value of the fixed slope
coef(summary(fit.Model.2))[4,1]
```

The standard deviation and autocorrelation of the level-1 residuals are
extracted as follows:

```{r chunk-summary-model-2-L1-Errors, echo = TRUE}
# Extract level-1 residuals standard deviation
as.numeric(VarCorr(fit.Model.2)[3,2])

# Extract level-1 residuals correlation between consecutive points
as.numeric(coef(fit.Model.2$modelStruct$corStruct,unconstrained=FALSE))
```

The standard deviation of the random intercept is given by:

```{r chunk-summary-model-2-RE-1, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random intercept
as.numeric(VarCorr(fit.Model.2)[1,2])
```

The standard deviation of the random slope is given by:

```{r chunk-summary-model-2-RE-2, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.2)[2,2])
```

The correlation between the random intercept and the random slope is given by:

```{r chunk-summary-model-2-RE-3, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.2)[2,3])
```

# Example 3. Estimate group differences in the autoregressive effect in multilevel AR(1) models

In this illustration, we are interested in estimating differences in the
autoregressive effect of Negative Affect between participants diagnosed with
major depressive disorder (MDD) and control subjects. The dataset contains 38
participants diagnosed with MDD and 40 control subjects.

First, for each individual, we are going to compute the lagged variable Negative
Affect. The variable Negative Affect is lagged within each day.

```{r chunk-lag-l1-predictor-PA, echo = TRUE}
# Create a lag variable: the data is lag within a person and days
NA.lag = rep(0,nrow(data))
subjno.i = unique(data$PID)

for (i in subjno.i) {
    n.i = which(data$PID == i)
    Day.i = data$day[n.i]

    for (t in unique(Day.i)) {
        k.i = n.i[which(data$day[n.i] == t)]
        NA.lag[k.i] = shift(data$NA.[k.i], 1)
    }
}

data = cbind(data,NA.lag)
```

The lagged variable `NA.lag` will be centered using the individual's mean.

```{r chunk-centering-lag-l1-predictor-PA, echo = TRUE}
# Centered within individuals PA.lag

N.i = unique(data$PID)
NA.lag.c = rep(0, nrow(data))

for(i in N.i) {
    NA.lag.c[which(data$PID==i)] = data$NA.lag[which(data$PID==i)] -
    mean(data$NA.[which(data$PID==i)],na.rm=TRUE)
}

data = cbind(data,NA.lag.c)
```

To estimate the model, we use the function `lme` from the `nlme` R package. The
dependent variable is the Negative Affect (i.e. NA.), the predictor is the
lagged outcome, which is centered using the individuals' mean:

```{r chunk-fit-model-3, echo = TRUE}
# fit a linear mixed-effects model to data
fit.Model.3 = lme(NA. ~ 1 + MDD + NA.lag + MDD*NA.lag, random = ~1 + NA.lag|PID, na.action=na.omit, data=data, method="REML")
```

where `NA.` is the negative affect, `1` is the fixed intercept, `MDD` is the difference in the fixed intercept between the two groups, `NA.lag.c` is the fixed autoregressive effect and `MDD*NA.lag.c` is the difference in the fixed autoregressive effect between the two groups. The random effect structure of the model is `1 + NA.lag.c|PID`, where `1` is the random intercept, and `NA.lag.c` is the random slope, which is allowed to vary over participants (`PID`).

The summary of the estimation results is given by:

```{r}
summary(fit.Model.3)
```

We extract the estimated fixed intercept as follows,

```{r chunk-summary-model-3-FE, echo = TRUE}
# Extract fixed effect coefficients
# extract the value of fixed intercept
coef(summary(fit.Model.3))[1,1]
```

the differences on the intercept between the two groups is given by:

```{r chunk-summary-model-3-FE-1, echo = TRUE}
# extract the value of the difference in the fixed intercept between the two groups
coef(summary(fit.Model.3))[2,1]
```

the fixed autoregressive effect is:

```{r chunk-summary-model-3-FE-2, echo = TRUE}
# extract the value of fixed slope
coef(summary(fit.Model.3))[3,1]
```

and the difference in the autoregressive effect between the two groups is extracted as follows:

```{r chunk-summary-model-3-FE-3, echo = TRUE}
# extract the value of the difference in the fixed slope between the two groups
coef(summary(fit.Model.3))[4,1]
```

The standard deviation of the level-1 residuals is extracted as follows:

```{r chunk-summary-model-3-L1-Errors, echo = TRUE}
# Extract level-1 residuals standard deviation
as.numeric(VarCorr(fit.Model.3)[3,2])
```

The standard deviation of the random intercept is given by:

```{r chunk-summary-model-3-RE-1, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random intercept
as.numeric(VarCorr(fit.Model.3)[1,2])
```

The standard deviation of the random slope is given by:

```{r chunk-summary-model-3-RE-2, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.3)[2,2])
```

The correlation between the random intercept and the random slope is given by:

```{r chunk-summary-model-3-RE-3, echo = TRUE}
# Extract random effect covariance structure
# Extract the standard deviation of the random slope
as.numeric(VarCorr(fit.Model.3)[2,3])
```

# Get the session info (also for reproducibility)

```{r chunk-session-info, echo = TRUE}
sessionInfo()
```

## References
