---
title: Power Analysis for $\\text{AR}(1)$ Multilevel Models in Intensive Longitudinal Designs
author:
  - "[Ginette Lafit](/presenters/ginette-lafit.md)"
date: today
toc: true
format:
    html: default
    pdf: default
execute:
    freeze: auto
    eval: true
    echo: true
---

## Settings things up

Before we proceed, we need to ensure we have several packages installed and
loaded into our `R` session. For the scripts below, we will use the following
packages:

- [`htmltools`](https://CRAN.R-project.org/package=htmltools)
- [`DT`](https://CRAN.R-project.org/package=DT)
- [`nlme`](https://CRAN.R-project.org/package=nlme)
- [`ggplot2`](https://CRAN.R-project.org/package=ggplot2)
- [`gridExtra`](https://CRAN.R-project.org/package=gridExtra)
- [`data.table`](https://CRAN.R-project.org/package=data.ta)
- [`plyr`](https://CRAN.R-project.org/package=plyr)
- [`dplyr`](https://CRAN.R-project.org/package=dplyr)
- [`formattable`](https://CRAN.R-project.org/package=formattable)
- [`tidyr`](https://CRAN.R-project.org/package=tidyr)
- [`MASS`](https://CRAN.R-project.org/package=MASS)
- [`compiler`](https://CRAN.R-project.org/package=compiler)
- [`kableExtra`](https://CRAN.R-project.org/package=kableExtra)
- [`future.apply`](https://CRAN.R-project.org/package=future.ap)
- [`PowerAnalysisIL`](https://CRAN.R-project.org/package=PowerAnalysisIL)
- [`tictoc`](https://CRAN.R-project.org/package=tictoc)
- [`viridis`](https://CRAN.R-project.org/package=viridis)
- [`ggpubr`](https://CRAN.R-project.org/package=ggpubr)

Which we can install in one go as follows:

```{r}
#| name: install-packages
#| echo: true
#| eval: false
#| warning: false

# Prepare the package list.
packages = c(
    "htmltools", "DT", "nlme", "ggplot2", "gridExtra",
    "data.table", "plyr", "dplyr", "formattable", "tidyr",
    "MASS", "compiler", "kableExtra", "future.apply",
    "tictoc", "viridis", "ggpubr", "devtools"
)

# Install packages.
install.packages(packages)
```

::: {.callout-tip}
You may consider first checking if the packages are installed before
actually installing them. Nevertheless, the code above will not reinstall
packages that are already installed and up-to-date.
:::

At last, we can load the packages into our `R` session:

```{r}
#| name: load-packages
#| echo: true
#| eval: false
#| warning: false

library(htmltools)
library(DT)
library(nlme)
library(ggplot2)
library(gridExtra)
library(data.table)
library(plyr)
library(dplyr)
library(formattable)
library(tidyr)
library(MASS)
library(compiler)
library(future.apply)
library(tictoc)
library(kableExtra)
library(viridis)
library(ggpubr)
```

Finally, we will also install the `Shiny` application package `PowerAnalysisIL`
because it contains the functions we will use to conduct the power analysis.

```{r}
# Install the shiny application package.
devtools::install_github("ginettelafit/PowerAnalysisIL")

## Load package `PowerAnalysisIL`.
library(PowerAnalysisIL)
```

## Description

We now focus on conducting a **sensitivity** power analysis to investigate
different factors that influence statistical power when testing group
differences in the autoregressive effect.

The goal of the new study is to determine the sample size (i.e., $N_0$ the
number of control participants and $N_1$ the number of participants diagnosed
with depression), when assuming the new study will include 70 repeated
measurement occasions.

To estimate this research question we will use a multilevel $\text{AR}(1)$ model
with a cross-level interaction effect between a level 1 continuous predictor
(lagged NA) and level 2 binary predictor (diagnosis):

\begin{multline}
    \text{NA}_{it} = \beta_{00} + \beta_{01} \text{Diagnosis}_{i} + \beta_{10} \text{NA}_{it-1} + \\
    \beta_{01} \text{Diagnosis}_{i}\text{NA}_{it-1} + \nu_{0i} + \nu_{1i} \text{NA}_{it-1} + \epsilon_{it}
\end{multline}

where $\beta_{00}$ is the fixed intercept, $\beta_{01}$ represents the main
effect of Diagnosis, $\beta_{10}$ is the fixed autoregressive effect, and
$\beta_{11}$ represents the differences in the autoregressive effect between
persons diagnosed with depression and control participants. The level 1
predictor (lagged NA) is centered within-persons and within-days.

In the model above, we assume the level 1 errors are independent and normally
distributed with mean zero and variance $\sigma_\epsilon^2$.

Between-person differences in the autoregressive effect are captured by
including a random intercept $\nu_{0i}$ and random slope $\nu_{1i}$. These
random effects are multivariate normal distributed with mean zero and covariance
matrix $\Sigma_\nu$:

$$
\boldsymbol{\Sigma}_\nu = \begin{bmatrix}
\sigma_{\nu_{0}}^2 & \sigma_{\nu_{01}}  \\
\sigma_{\nu_{01}} & \sigma_{\nu_{1}}^2
\end{bmatrix}
$$

In this model, it is also assumed that the level 2 random effects and the level
1 errors are independent.

To investigate group differences in the autoregressive effect of NA between
persons diagnosed with depression and control participants we are going to
conduct the following hypothesis test:

$$
H_0: \beta_{11} = 0
$$

$$
H_1: \beta_{11} \neq 0
$$

The goal of this exercise is to conduct a sensitivity power analysis to
investigate:

- differences in statistical power when varying the number of measurement
  occasions $T$ due to different levels of compliance (i.e., 60\% and 80\%)
- differences in statistical power when varying the value of $\beta_{11}$: we
  assume $\beta_11$ is 10\% lower/higher than the one obtained using the Leuven
  clinical data set

## Determining model parameter values

To obtain the values of the model parameters we will use data from the Leuven
clinical study. The code to estimate the multilevel model is included in the exercise
[`Multilevel model estimation using the Leuven Clinical Dataset`](estimation-multilevel-leuven-clinical-data.Rmd).
The output of the fitted model is:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/fit-multilevel-model-example-3.png)

Thus, the values of the model parameter that will be used to conduct the power
analysis are:

$$
\begin{aligned}
    \beta_{00} &= 6.82 \; \; \; \text{fixed intercept} \\
    \beta_{01} &= 16.33 \; \; \; \text{main effect of diagnosis} \\
    \beta_{10} &= 0.31 \; \; \; \text{fixed autoregressive effect} \\
    \beta_{10} &= 0.12 \; \; \; \text{cross-level interaction effect} \\
    \sigma_\epsilon &= 8.75 \; \; \; \text{std. deviation level 1 errors} \\
    \sigma_{\nu_0} &= 5.79 \; \; \; \text{std. deviation random intercept} \\
    \sigma_{\nu_1} &= 0.14 \; \; \; \text{std. deviation random slope} \\
    \rho_{\nu_{01}} &= -0.199 \; \; \; \text{correlation between the random effects}
\end{aligned}
$$

## Set the model parameters values

For our sensitivity power analysis, we can use the estimated parameter values
using the Leuven clinical data set:

```{r}
#| name: model-parameters
#| warning: false

# Number of persons in reference group (Control participants).
N.0 = c(20, 40, 60, 100)

# Number of participants diagnosed with MDD.
N.1 = c(20, 40, 60, 100)

# Number of measurement occasions.
T.obs = 70

# If Ylag.center = TRUE, person-mean centered the lagged outcome, otherwise the lagged predictor is not person-mean centered.
Ylag.center = FALSE

# Fixed intercept for participants in the reference group.
b00 = 6.82

# Differences in the fixed intercept between the two groups.
b01.Z = 16.33

# Fixed autoregressive effect for participants in the reference group.
b10 = 0.31

# Differences in the fixed autoregressive effect between the two groups.
b11.Z = 0.12

# Std. deviation of the level 1 errors.
sigma = 8.75

# Std. deviation of the random intercept.
sigma.v0 = 5.79

# Std. deviation of the random autoregressive effect.
sigma.v1 = 0.14

# Correlation between the random intercept and random autoregressive effect.
rho.v = -0.199

# Significant level.
alpha = 0.05

# Select the side of the test as follows.

# One-side test H1: b01.Z > 0
# side.test = 1

# One-side test H1: b01.Z < 0
# side.test = 2

# One-side test H1: b01.Z is different from 0
side.test = 3

# Set the optimization method for `lme4`.
Opt.Method = "REML"

# Number of Monte Carlo replicates.
R = 10

# Set seed or the random number generator for reproducibility.
set.seed(123)
```

::: {.callout-important}
In the code above we set the number of replications `R` to $10$. This is just so
we can compile the document quickly. In practice, you should set `R` to a much
higher number. We recommend at least $1000$ replications.
:::

## Conduct the sensitivity power analysis

We can now conduct a sensitivity analysis to assess how power varies under
different conditions. First, we focus on the scenario in which we varied the
value of the main effect of interest.

### Varying the value of the main effect of interest

```{r}
#| name: estimate-power-sensitivity-1
#| echo: true
#| eval: false
#| warning: false

# We investigate what happens with power when the main effect of interest is 10%
# larger/smaller than the expected value based on previous studies.

# Differences in the fixed autoregressive effect between the two groups.
b11.Z.list <- b11.Z * c(0.9, 1, 1.10)

# Prepare the lists where we will store the results.
Power.Simulation.list.REML.list <- list()
R.Converge.Simulation.list.REML.list <- list()

# Function for computing power using simulation-based approach using REML.
for (i in 1:length(b11.Z.list)) {
    Power.Simulation.list.REML.list[[i]] <- lapply(1:length(N.0), function(n) {
        Power.Curve.Simulation.ML.AR.Group.Diff(
            N.0[n],
            N.1[n],
            T.obs,
            Ylag.center,
            b00,
            b01.Z,
            b10,
            b11.Z.list[i],
            sigma,
            sigma.v0,
            sigma.v1,
            rho.v,
            alpha,
            side.test,
            Opt.Method = "REML",
            R
        )
    })

    # Number of replicates that converge.
    R.Converge.Simulation.list.REML.list[[i]] <- unlist(
        lapply(1:length(N.0), function(n) {
            Power.Simulation.list.REML.list[[i]][[n]]$n.R
        })
    )
}

# Print the number of replicates that converged.
R.Converge.Simulation.list.REML.list
```

We plot the power curves for the sensitivity analysis for the value of the
difference in NA between participants diagnosed with MDD and participants in the
control group.

```{r}
#| name: estimate-power-effect-size
#| echo: true
#| eval: false
#| warning: false

# Construct a matrix with the results of the sensitivity power analysis.
Power.b11.power <- matrix(0, length(N.0), length(b11.Z.list))

# Get the values of power.
for (i in 1:length(b11.Z.list)) {
    Power.b11.power[, i] = unlist(
        lapply(
            1:length(N.0), function(n) {
                Power.Simulation.list.REML.list[[i]][[n]]$power.hat.lme["b11.Z"]
            }
        )
    )
}

# Convert to long format.
Power.b11.power = melt(Power.b11.power, id.vars = "Power")

# Add column names.
colnames(Power.b11.power) <- c("N.0", "Value", "Power")

# Store the values.
Power.b11.power$N.0 <- rep(N.0, length(b11.Z.list))
Power.b11.power$N.1 <- rep(N.1, length(b11.Z.list))
Power.b11.power$Value <- rep(b11.Z.list, each = length(N.0))
Power.b11.power <- Power.b11.power[, c("N.0", "N.1", "Value", "Power")]

# Convert to data frame.
Power.b11.power <- data.frame(Power.b11.power)

# Create curve with the results of the sensitivity power analysis.
ggplot(Power.b11.power, aes(
        x = N.0,
        y = Power,
        colour = as.factor(Value),
        group = as.factor(Value)
    )) +
    geom_line(
        size = 1
    ) +
    geom_point(
        size = 1
    ) +
    labs(
        title = "Power curves for the difference in
                the inertia of NA between participants \n
                diagnosed with MDD and participants in the
                control group",
        x = "Number of Persons",
        color = "Difference in \n
                inertia of NA"
    ) +
    scale_x_continuous(
        name = "Participants",
        breaks = N.0,
        labels = c(paste(N.0, N.1, sep = ";"))
    ) +
    theme(
        axis.text.x = element_text(angle = 90)
    ) +
    scale_colour_manual(
        values = c("#CC79A7", "#56B4E9", "#009E73")
    ) +
    geom_hline(
        yintercept = 0.9, linetype = "dashed"
    ) +
    theme(
        legend.position = "bottom",
        legend.title = element_text(size = 10)
    ) +
    theme_minimal()
```

We now conduct a sensitivity analysis to assess how power varies under different
conditions. This time, we focus on the scenario in which we varied the average
compliance rate.

[Ginette, Can you please take a look at the section above and the
following?]{.todo}

### Varying the average compliance rate

```{r}
#| name: estimate-power-sensitivity-2
#| echo: true
#| eval: false
#| warning: false

# Average compliance rate.
T.obs.list <- T.obs * c(0.6, 0.8, 1)

# Prepare the lists where we will store the results.
Power.Simulation.list.REML.Tpoints.list <- list()
R.Converge.Simulation.list.REML.Tpoints.list <- list()

# Function for computing power using simulation-based approach using REML.
for (i in 1:length(T.obs.list)) {
    Power.Simulation.list.REML.Tpoints.list[[i]] <- lapply(1:length(N.0), function(n) {
        Power.Curve.Simulation.ML.AR.Group.Diff(
            N.0[n],
            N.1[n],
            T.obs.list[i],
            Ylag.center,
            b00,
            b01.Z,
            b10,
            b11.Z,
            sigma,
            sigma.v0,
            sigma.v1,
            rho.v,
            alpha,
            side.test,
            Opt.Method = "REML",
            R
        )
    })

    # Number of replicates that converge.
    R.Converge.Simulation.list.REML.Tpoints.list[[i]] <- unlist(
        lapply(
            1:length(N.0), function(n) {
                Power.Simulation.list.REML.Tpoints.list[[i]][[n]]$n.R
            }
        )
    )
}

# Number of replicates that converge.
R.Converge.Simulation.list.REML.Tpoints.list
```

We plot the power curves for the sensitivity analysis for the value of the
difference in NA between participants diagnosed with MDD and participants in the
control group.

```{r}
#| name: estimate-power-time-point
#| echo: true
#| eval: false
#| warning: false

# Construct a matrix with the results of the sensitivity power analysis.
Power.b11.power.Tpoints <- matrix(0, length(N.0), length(T.obs.list))

# Get the values of power.
for (i in 1:length(T.obs.list)) {
    Power.b11.power.Tpoints[, i] <- unlist(
        lapply(
            1:length(N.0), function(n) {
                Power.Simulation.list.REML.Tpoints.list[[i]][[n]]$power.hat.lme["b11.Z"]
            }
        )
    )
}

# Transform to format for the plot.
Power.b11.power.Tpoints <- melt(Power.b11.power.Tpoints, id.vars = "Power")

# Add column names.
colnames(Power.b11.power.Tpoints) <- c("N.0", "Value", "Power")

# Store the values.
Power.b11.power.Tpoints$N.0 <- rep(N.0, length(T.obs.list))
Power.b11.power.Tpoints$N.1 <- rep(N.1, length(T.obs.list))
Power.b11.power.Tpoints$Value <- rep(T.obs.list, each = length(N.0))
Power.b11.power.Tpoints <- Power.b11.power.Tpoints[, c("N.0", "N.1", "Value", "Power")]

# Create curve with the results of the sensitivity power analysis.
Power.b11.power.Tpoints <- data.frame(Power.b11.power.Tpoints)

# Create curve with the results of the sensitivity power analysis.
ggplot(Power.b11.power.Tpoints, aes(
        x = N.0,
        y = Power,
        colour = as.factor(Value),
        group = as.factor(Value)
    )) +
    geom_line(
        size = 1
    ) +
    geom_point(
        size = 1
    ) +
    labs(
        title = "Power curves for the difference in
                the inertia of NA between participants \n
                diagnosed with MDD  and participants in
                the control group",
        x = "Number of Persons",
        color = "Number of \n time points"
    ) +
    scale_x_continuous(
        name = "Participants",
        breaks = N.0,
        labels = c(paste(N.0, N.1, sep = ";"))
    ) +
    theme(
        axis.text.x = element_text(angle = 90)
    ) +
    scale_colour_manual(
        values = c("#CC79A7", "#56B4E9", "#009E73")
    ) +
    geom_hline(
        yintercept = 0.9,
        linetype = "dashed"
    ) +
    theme(
        legend.position = "bottom",
        legend.title = element_text(size = 10)
    ) +
    theme_minimal()
```

## Session information

Using the command below, we can print the `session` information (i.e., operating
system, details about the `R` installation, and so on) for reproducibility
purposes.

:::{.code-font-small}
```{r}
#| name: session-info
#| warning: false

# Session information.
sessionInfo()
```
:::
