---
title: "Power Analysis for Multilevel Models With Cross-level Interactions in Intensive Longitudinal Designs"
author:
  - "[Ginette Lafit](/presenters/ginette-lafit.md)"
date: today
toc: true
format:
    html: default
    pdf: default
execute:
    freeze: auto
    eval: true
    echo: true
---

## Setting things up

Before we proceed, we need to ensure we have several packages installed and
loaded into our `R` session. For the scripts below, we will use the following
packages:

- [`tidyverse`](https://CRAN.R-project.org/package=tidyverse)
- [`gridExtra`](https://CRAN.R-project.org/package=gridExtra)
- [`formattable`](https://CRAN.R-project.org/package=formattable)
- [`htmltools`](https://CRAN.R-project.org/package=htmltools)
- [`shiny`](https://CRAN.R-project.org/package=shiny)
- [`DT`](https://CRAN.R-project.org/package=DT)
- [`ggplot2`](https://CRAN.R-project.org/package=ggplot2)
- [`plyr`](https://CRAN.R-project.org/package=plyr)
- [`dplyr`](https://CRAN.R-project.org/package=dplyr)
- [`tidyr`](https://CRAN.R-project.org/package=tidyr)
- [`shinyjs`](https://CRAN.R-project.org/package=shinyjs)
- [`shinythemes`](https://CRAN.R-project.org/package=shinythemes)
- [`viridis`](https://CRAN.R-project.org/package=viridis)
- [`plotly`](https://CRAN.R-project.org/package=plotly)
- [`remotes`](https://CRAN.R-project.org/package=remotes)
- [`nlme`](https://CRAN.R-project.org/package=nlme)
- [`devtools`](https://CRAN.R-project.org/package=devtools)
- [`data.table`](https://CRAN.R-project.org/package=data.table)
- [`MASS`](https://CRAN.R-project.org/package=MASS)
- [`compiler`](https://CRAN.R-project.org/package=compiler)
- [`future.apply`](https://CRAN.R-project.org/package=future.apply)

Which we can install in one go as follows:

```{r}
#| name: install-packages
#| echo: true
#| eval: false
#| warning: false

# Prepare the package list.
packages = c(
    "tidyverse", "gridExtra", "formattable", "htmltools", "shiny",
    "DT", "ggplot2", "plyr", "dplyr", "tidyr", "shinyjs", "shinythemes",
    "viridis", "plotly", "remotes", "nlme", "devtools", "data.table",
    "MASS", "compiler", "future.apply"
)

# Install packages.
install.packages(packages)
```

::: {.callout-tip}
You may consider first checking if the packages are installed before
actually installing them. Nevertheless, the code above will not reinstall
packages that are already installed and up-to-date.
:::

At last, we can load the packages into our `R` session:

```{r}
#| name: load-packages
#| echo: true
#| eval: false
#| warning: false

# Load packages.
library(tidyverse)
library(gridExtra)
library(formattable)
library(htmltools)
library(shiny)
library(DT)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(shinyjs)
library(shinythemes)
library(viridis)
library(plotly)
library(remotes)
library(nlme)
library(devtools)
library(data.table)
library(MASS)
library(compiler)
library(future.apply)
```

## Description

The goal of this exercise is to conduct a power analysis to select the number of
persons to investigate if depression moderates the relationship between
anhedonia and negative affect (NA).

To estimate this research question we will use a multilevel model with a
cross-level interaction effect between a level 1 continuous predictor
(anhedonia) and level 2 continuos predictor (depression):

\begin{multline}
    \text{NA}_{it} = \beta_{00} + \beta_{01} \text{Depression}_{i} + \beta_{10} \text{Anhedonia}_{it} +\\
        \beta_{01} \text{Depression}_{i}\text{Anhedonia}_{it} +
        \nu_{0i} + \nu_{1i} \text{Anhedonia}_{it}  + \epsilon_{it}
\end{multline}

where $\beta_{00}$ is the fixed intercept, $\beta_{01}$ represents the main
effect of *depression*, $\beta_{10}$ is the fixed slope of *anhedonia*, and
$\beta_{11}$ represents the cross-level interaction effect between *anhedonia*
and *depression*. The cross-level interaction effect assesses whether
*depression* moderates the effect of *anhedonia* on NA. The level 1 predictor
(i.e., *anhedonia*) is centered within-persons and within-days.

In this model, we account for the serial dependency that characterizes IL
designs by assuming that the level 1 errors follow an autoregressive
$\text{AR}(1)$ process [see @goldstein1994multilevel]:

$$
\epsilon_{it} = \rho_\epsilon \epsilon_{it-1} + \varepsilon_{it}
$$

where the correlation between two consecutive errors is denoted by
$\rho_\epsilon$, and $\varepsilon_{it}$ is a Gaussian error with mean zero and
variance $\sigma_\epsilon^2$. Under this model, the variance of the level 1
errors is given by $\sigma_\epsilon^2/(1-\rho_\epsilon^2)$. To guarantee that
this model is stationary, the autocorrelation $\rho_\epsilon$ should range
between -1 and 1 [@hamilton1994time].

Between-person differences in the relation between *anhedonia* and NA are captured
by including a random intercept $\nu_{0i}$ and random slope $\nu_{1i}$. These
random effects are multivariate normal distributed with mean zero and covariance
matrix $\Sigma_\nu$:

$$
\boldsymbol{\Sigma}_\nu = \begin{bmatrix}
\sigma_{\nu_{0}}^2 & \sigma_{\nu_{01}}  \\
\sigma_{\nu_{01}} & \sigma_{\nu_{1}}^2
\end{bmatrix}
$$

In this model, it is also assumed that the level 2 random effects and the Level
1 errors are independent.

To investigate if depression moderates the relationship between anhedonia and NA
we are going to conduct the following hypothesis test:

$$
H_0: \beta_{11} = 0
$$

$$
H_1: \beta_{11} \neq 0
$$

The aim of this exercise is to select the number of participants assuming the
number of repeated measurements occasions to $T = 70$.

- Select sample size using the analytic approach, e.g., $N = \{20, 40, ...\}$.
- Compare the results with the ones obtained using the simulation-based approach.

## Determining model parameter values

To obtain the values of the model parameters we will use data from the Leuven
clinical study. The code to estimate the multilevel model with cross-level
interaction effect is included in the exercise
[`Multilevel model estimation using the Leuven Clinical Dataset`](estimation-multilevel-leuven-clinical-data.Rmd).
The output of the fitted model is:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/fit-multilevel-model-example-2.png)

Thus, the values of the model parameter that will be used to conduct the power
analysis are:

$$
\begin{aligned}
    \beta_{00} &= 42.98 \; \; \; \text{fixed intercept} \\
    \beta_{01} &= 1.53 \; \; \; \text{main effect of depression} \\
    \beta_{10} &= 0.14 \; \; \; \text{fixed slope} \\
    \beta_{10} &= -0.01 \; \; \; \text{cross-level interaction effect} \\
    \sigma_\epsilon &= 11.92 \; \; \; \text{std. deviation level 1 errors} \\
    \rho_\epsilon &= 0.43 \; \; \; \text{std. deviation level 1 errors} \\
    \sigma_{\nu_0} &= 12.86 \; \; \; \text{std. deviation random intercept} \\
    \sigma_{\nu_1} &= 0.11 \; \; \; \text{std. deviation random slope} \\
    \rho_{\nu_{01}} &= 0.249 \; \; \; \text{correlation between the random effects} \\
    \mu_\text{Anhedonia} &= 51.66 \; \; \; \text{mean anhedonia} \\
    \sigma_\text{Anhedonia} &= 23.67 \; \; \; \text{std. deviation anhedonia} \\
    \mu_\text{Depression} &= 15.71 \; \; \; \text{mean anhedonia} \\
    \sigma_\text{Depression} &= 5.00 \; \; \; \text{std. deviation anhedonia}
\end{aligned}
$$

## Analytical-based power analysis

To conduct the power analysis using the analytic approach we are going to use
`ApproxPowerIL`: a `Shiny` application and `R` package to perform power analysis
to select the number of persons for multilevel models with auto-correlated
errors using asymptotic approximations of the information matrix The repository
contains functions used in @lafit2023enabling. Users can download the app and
run locally on their computer by executing the following commands in `R` or
`RStudio` at
[ApproxPowerIL](https://gitlab.kuleuven.be/ppw-okpiv/researchers/u0119584/ApproxPowerIL).

```{r}
#| name: approx-power-il-app
#| echo: true
#| eval: false
#| warning: false

# Install the app package from the `GitHub` repository.
remotes::install_github("ginettelafit/ApproxPowerIL", force = TRUE)

# Load package `ApproxPowerIL`.
library(ApproxPowerIL)

# Lunch the app from the `GitHub` gist.
shiny::runGist('302737dc046b89b7f09d15843389161c')
```

### Step 1

**Select the model and set the sample size in the `ApproxPowerIL` application.**

- Indicate the model of interest.
- Input the number of participants $N$ (comma-separated): $N = \{ 20, 40, 60, 80, 100 \}$.
- Input the number of repeated measurement occasions: $T = 70$.

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-approx-il-ex-1-1.png){width=60%}

### Step 2

**Set the value of the model parameters in the `ApproxPowerIL` application.**

Use the screenshots below to guide you through the process:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-approx-il-ex-1-2.png){width=60%}

And, for the remainder of the parameters:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-approx-il-ex-1-3.png){width=60%}


### Step 3

**Inspect the results.**

Statistical power is higher than 90% when the number of participants is equal to
or higher than $20$.

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-approx-il-ex-1-4.png)

## Simulation-based power analysis

To conduct the power analysis using the simulation-based approach we are going
to use `PowerAnalysisIL`: a `Shiny` application and `R` package to perform power
analysis to select the number of persons for multilevel models using the
simulation-based approach.

download the app and run locally on their computer by executing the following
The repository contains functions used in @lafit2021selection. Users can
commands in `R` or `RStudio` at
[PowerAnalysisIL](https://github.com/ginettelafit/PowerAnalysisIL).

```{r}
#| name: simulation-power-il-app
#| echo: true
#| eval: false
#| warning: false

# Install the app package from the `GitHub` repository.
library(devtools)
devtools::install_github("ginettelafit/PowerAnalysisIL", force = TRUE)

## Load package `PowerAnalysisIL`.
library(PowerAnalysisIL)

# Lunch the app from the `GitHub` gist.
shiny::runGist('6bac9d35c2521cc4fd91ce4b82490236')
```

### Step 1

**Select the model and set the sample size in the `PowerAnalysisIL` application.**

- Indicate the model of interest.
- Input the number of participants $N$ (comma-separated): $N = \{ 20, 40, 60, 80, 100 \}$.
- Input the number of repeated measurement occasions: $T = 70$.

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-analysis-il-ex-1-1.png){width=60%}

### Step 2

**Set the value of the model parameters in the `PowerAnalysisIL` application.**

Use the screenshots below to guide you through the process:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-analysis-il-ex-1-2.png){width=60%}

And, for the remainder of the parameters:

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-analysis-il-ex-1-3.png){width=60%}

### Step 3

**Inspect the results.**

Statistical power is higher than 90% when the number of participants is equal to
or higher than $20$.

![[Ginette, can you please add the figure caption here?]{.todo}](../assets/images/power-analysis-il-ex-1-4.png)

## Session information

Using the command below, we can print the `session` information (i.e., operating
system, details about the `R` installation, and so on) for reproducibility
purposes.

:::{.code-font-small}
```{r}
#| name: session-info
#| warning: false

# Session information.
sessionInfo()
```
:::

## References
